---
title: "St558HW5"
format: html
editor: visual
---

# Annie DiFrank Homework 5

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# Use pacman to check whether packages are installed, if not load them
if (!require("pacman")) install.packages("pacman")
library(pacman)
pacman::p_load(dplyr,
 tidyverse,
 purrr,
 ggplot2,
 psych)

```

## Task 1: Reading in and modifying data

-   First, will load the data in locally

-   Second, will load and combine the data using functions from tidyverse

-   Lastly, will convert interesting categorical variables to factors for both math data, Portuguese, and combined data. The four variables I am interested in are

    -   address: student's home address type (binary: 'U' - urban or 'R' - rural)

    -   famsize: family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)

    -   Pstatus: parent's cohabitation status (binary: 'T' - living together or 'A' - apart)

    -   guardian: student's guardian (nominal: 'mother', 'father', or 'other')

```{r task1}

#Loading in data from local file sourse and combining the 2 dataframes
math_data = read.table("student-mat.csv", sep = ";", header = TRUE)
por_data = read.table("student-por.csv", sep = ";", header = TRUE)

combined_data = merge(
  math_data,
  por_data,
  by = c(
    "school",
    "sex",
    "age",
    "address",
    "famsize",
    "Pstatus",
    "Medu",
    "Fedu",
    "Mjob",
    "Fjob",
    "reason",
    "nursery",
    "internet"
  )
)
print(nrow(combined_data)) # 382 students

#Loading in the data and combing using functions from the tidyverse

math_data2 = read_csv2("student-mat.csv", col_names = TRUE)
por_data2 = read_csv2("student-por.csv", col_names = TRUE)

combine_data2 <-
  inner_join(
    math_data2,
    por_data2,
    by = c(
      "school",
      "sex",
      "age",
      "address",
      "famsize",
      "Pstatus",
      "Medu",
      "Fedu",
      "Mjob",
      "Fjob",
      "reason",
      "nursery",
      "internet"
    )
  )
#Warning: Detected an unexpected many-to-many relationship between `x` and `y`. Looked up this warning and it says it arises when a single row in one dataset can match with multiple rows in the other and vice versa, creating a "many to many" relationship. This can cause duplicate rows or incorrect data combinations - issue is with G1-3 and paid and absences because they are specific to the class

#grabbing variable names to copy paste
colnames(math_data)

comb_data3 <-
  inner_join(
    math_data2,
    por_data2,
    by = c(
      "school",
      "sex",
      "age",
      "address",
      "famsize",
      "Pstatus",
      "Medu",
      "Fedu",
      "Mjob",
      "Fjob",
      "reason",
      "guardian",
      "traveltime",
      "studytime",
      "failures",
      "schoolsup",
      "famsup",
      "activities",
      "nursery",
      "higher",
      "internet",
      "romantic",
      "famrel",
      "freetime",
      "goout",
      "Dalc",
      "Walc",
      "health"
    )
  )

#Converting 4 categorical variables to factor variables in each tibble using mutate()

math_data2 %>% mutate(
  "famsizeF" = as.factor(famsize),
  "PstatusF" = as.factor(Pstatus),
  "guardianF" = as.factor(guardian),
  "addressF" = as.factor(address),
  .keep = "unused"
) -> math_data3

por_data2 %>% mutate(
  "famsizeF" = as.factor(famsize),
  "PstatusF" = as.factor(Pstatus),
  "guardianF" = as.factor(guardian),
  "addressF" = as.factor(address),
  .keep = "unused"
) -> por_data3

comb_data3 %>% mutate(
  "famsizeF" = as.factor(famsize),
  "PstatusF" = as.factor(Pstatus),
 "guardianF" = as.factor(guardian),
 "addressF" = as.factor(address),
 .keep = "unused"
) -> comb_data3

#printing for display purposes
math_data3
por_data3
comb_data3

```

## Task 2: Summarizing the data

Here, we will look at how the data is stored to make sure everything makes sense as well as document any missing values. Then, we will perform summaries and graphs on interesting things.

NOTE: With doing psych::describe, I did not use (head) to look at the data- I added it in after the fact to de-clutter the output quarto document a little for grading ease.

### Part 1: EDA- clean up

```{r t2EDA}

#comb_data3

str(comb_data3)

#converting some dbl types to characters (categorical vars)
comb_data3 %>% mutate(
  "Medu" = as.character(Medu),
  "Fedu" = as.character(Fedu),
  "traveltime" = as.character(traveltime),
  "studytime" = as.character(studytime),
  "failures" = as.character(failures),
  "famrel" = as.character(famrel),
  "freetime" = as.character(freetime),
  "goout" = as.character(goout),
  "Dalc" = as.character(Dalc),
  "Walc" = as.character(Walc),
  "health" = as.character(health),
  .keep = "unused"
) -> comb_data3

head(psych::describe(comb_data3))#nothing looks weird here

colSums(is.na(comb_data3)) #no NAs which is expected, says in their documentation

#Im going to rename the Gs/ paid/abscences for my own benefit
comb_data3 %>% mutate(
  "G1.math" = G1.x,
  "G1.por" = G1.y,
  "G2.math" = G2.x,
  "G2.por" = G2.y,
  "G3.math" = G3.x,
  "G3.por" = G3.y,
  "paid.math" = paid.x,
  "absences.math" = absences.x,
  "paid.por" = paid.y,
  "absences.por" = absences.y,
  .keep = "unused"
) -> comb_data3

```

### Task 2 Part 2: Categorical Variables

```{r t2CATVARS}

#one-way table
table(comb_data3$famsizeF)
#Over double the amount of people with less than or equal to 3 members have >3 family members; 230 have >3 and 90 have <= 3.

#two-way table
table(comb_data3$famsizeF, comb_data3$addressF)
#Regardless of family size, urban living is more common than rural. Out of the people with families >3 people, 53 live rurally.

#three-way table
table(comb_data3$famsizeF,
      comb_data3$addressF,
      comb_data3$PstatusF)
#First observation, significantly more people have parents living together which is nice. Of the people with split parents, 3/16 of those with families >3 people live rurally.
#____________________________________________________

#conditional two-way table

threeway <-
  table(comb_data3$famsizeF,
        comb_data3$addressF,
        comb_data3$PstatusF)
threeway[, , 2]
#This gives only the values of family size and address for those with parents that are together
comb_data3 %>% filter(PstatusF == "T") -> subsetted_math
table(subsetted_math$famsizeF, subsetted_math$addressF)
#This gives only the values of family size and address for those with parents that are together
#Out of the people with parents together, 12 of the people with less than or equal to 3 family size live rurally.

#_____________________________________________________

#Two-way contigency table 

comb_data3 %>% group_by(famsizeF, addressF) %>% summarize(count = n()) %>% pivot_wider(names_from =
                                                                                         addressF, values_from = count) #same output as the two-way table from earlier!

#______________________________________________________

#Stacked bar plot
ggplot(comb_data3, aes(x = famsizeF, fill = addressF)) + geom_bar() + labs(x = "Family size: >3 or <=3", y = "Number of participants", title = "Distribution of address with family size among participants") + scale_fill_discrete("Address: Rural or Urban")

#Side-by-side plot
ggplot(comb_data3, aes(x = famsizeF, fill = addressF)) + geom_bar(position = "dodge") + labs(x = "Family size: >3 or <=3", y = "Number of participants", title = "Distribution of address with family size among participants") + scale_fill_discrete("Address: Rural or Urban")

#Both graphs show that regardless of family size, there are significantly more participants living urbally than rurally. If I ran a statistically test I would bet there's no significant interaction between the two. 
```

### Task 2 Part 3: Numeric variables (and across groups)

Numeric variables: age, absences, and the three test grade variables from each data set (math and Portuguese).

Here we will:

-    find the center and spread for age, absences of the math class, and G1 of the math class.

    -   Repeat while subsetting the data in a meaningful way.

-    Find measures of center and spread across a single grouping variable for three of these variables
    (including a G3 variable as one of them)

-    Find measures of center and spread across two grouping variables for three of these variables

-   Create a correlation matrix between all of the numeric variables

```{r t2NUMVARSpt1}

#Center and spread for each variable
#Age
comb_data3 %>% summarize(mean = mean(age),
                         med = median(age),
                         stddev = sd(age)) -> age
#Absences of math class
comb_data3 %>% summarize(
  mean = mean(absences.math),
  med = median(absences.math),
  stddev = sd(absences.math)
) -> absences_math
#G1 of math class
comb_data3 %>% summarize(
  mean = mean(G1.math),
  med = median(G1.math),
  stddev = sd(G1.math)
) -> G1_math

NumSums <- bind_rows(age, absences.math, G1.math) %>% as.data.frame()
rownames(NumSums) = c("age", "absences_math", "G1_math")

#Repeating while subsetting data to just females
#Age
comb_data3 %>% filter(sex == "F") %>% summarize(mean = mean(age),
                                                med = median(age),
                                                stddev = sd(age)) -> Female_age
#Absences of math class
comb_data3 %>% filter(sex == "F") %>% summarize(
  mean = mean(absences.math),
  med = median(absences.math),
  stddev = sd(absences.math)
) -> Female_absences_math
#G1 of math class
comb_data3 %>% filter(sex == "F") %>% summarize(
  mean = mean(G1.math),
  med = median(G1.math),
  stddev = sd(G1.math)
) -> Female_G1_math
NumSumsF <-
  bind_rows(Female_age, Female_absences.math, Female_G1.math) %>% as.data.frame()
rownames(NumSumsF) = c("Fem_age", "Fem_absences_math", "Fem_G1_math")

NumSums <- bind_rows(NumSums, NumSumsF)
NumSums

#Looking at all the data, it looks like Absences, regardless of whether its whole class or just females, are very variable. Just females did slightly lower grade wise than both sexes combined. 

#_________________________________________________________
#Find measures of center and spread across a single grouping variable for three of these variables

comb_data3 %>% group_by(sex) %>% summarize(mean = mean(age),
                                           med = median(age),
                                           stddev = sd(age)) -> age_grouped

comb_data3 %>% group_by(sex) %>% summarize(
  mean = mean(absences.math),
  med = median(absences.math),
  stddev = sd(absences.math)
) -> absences_math_grouped

comb_data3 %>% group_by(sex) %>% summarize(
  mean = mean(G3.math),
  med = median(G3.math),
  stddev = sd(G3.math)
) -> G3_math_grouped

NumSums_Grouped <-
  bind_rows(age_grouped, absences_math_grouped, G3_math_grouped) %>% as.data.frame()
rownames(NumSums_Grouped) = c("ageF",
                              "ageM",
                              "absences_mathF",
                              "absences_mathM",
                              "G3F",
                              "G3M")
NumSums_Grouped

#It looks like females had a slightly higher amount of absences (but variance is very big on those counts) and males performed slightly better on G3. Age is about the same between sexes.

#________________________________________________________
#Correlation matrix between all numeric variables 
num_cols <- comb_data3 %>% select(where(is.numeric))

correlation_matrix <- cor(num_cols, use = "complete.obs")
correlation_matrix
#Noticeable, and probably expected, strong corrrelations are between the 3 G's. There's no strong correlations with age, absences between math and portuguese have a little correlation which makes sense if it's the same student missing both classes. 
```

Here we will do visuals:

-    Create a histogram, kernel density plot, and boxplot for two of the numeric variables across one of the categorical variables

-   Create two scatterplots relating a G3 variable to other numeric variables (put G3 on the y-axis).

-   Repeat the scatter plot step but use faceting to obtain graphs at each setting of another categorical
    variable.

-   Repeat the scatter plot step but use faceting to obtain graphs at each combination of two categorical
    variables

```{r t2NUMVARSpt2}

#Histogram of age and G3 across school 
ggplot(comb_data3, aes(x = age, fill = school)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  geom_histogram(aes(x = G3.math, fill = school),
                 alpha = 0.5,
                 position = "identity") +
  labs(x = "Age and Grade 3 of Math", fill = "School", title = "Age and Grade 3 of Math across School") +
  theme_minimal()
#There are way more students at GP than MS school. Students at MS had a little more of a linear trend in age, while GP school had more of a normal distribution.

#Kernel density plot of age and G3 across school
ggplot(comb_data3, aes(x = age, fill = school)) +
  geom_density(alpha = 0.5) +
  geom_density(aes(x = G3.math, fill = school),
               alpha = 0.5,
               linetype = "dashed") +
  labs(x = "Age and Grade 3 of Math", fill = "School", title = "Age and Grade 3 of Math across School") +
  theme_minimal()
#This plot is easier to see the seperation of numeric variables in than the histogram. Both schools have a pretty normal distribution of G3, while MS seems to have older students on average than GP.

#Boxplot of age and G3 across school

#reshaping data to long format 
comb_long <-
  comb_data3 %>% pivot_longer(
    cols = c(age, G3.math),
    names_to = "variable",
    values_to = "value"
  )

ggplot(comb_long, aes(x = variable, y = value, fill = school)) +
  geom_boxplot(alpha = 0.5) +
  labs(x = "Age and Grade 3 of Math",
       y = "Value",
       fill = "School",
       title = "Age and Grade 3 of Math across School") +
  theme_minimal()
  #GP on average looks to have higher Grade 3s than MS. Age is pretty similar with low variance, but the age is slightly higher on average at MS. The outliers in G3 are probably people that dropped out of the class. 

#----------------------------------------------------------------------------

#Scatterplot of G3 as a Y axis and age with sex as a fill 
ggplot(comb_data3, aes(x = age, y = G3.math, color = sex)) + geom_point() +
  geom_jitter(width = 0.35, opacity = 0.3) + labs(x = "Age", y = "Grade 3 of Math Course", title = "Grade 3 of math by age, filled by sex")
#There doesnt seem to be any difference in grades between sexes. There are two outliers, two males are 22 years old.

#Scatterplot of G3 as a Y axis and G1 with guardian as a fill
ggplot(comb_data3, aes(x = G1.math, y = G3.math, color = guardianF)) + geom_point() +
  geom_jitter(width = 0.35, opacity = 0.3) + labs(x = "Grade 1 of Math Course", y = "Grade 3 of Math Course", title = "Grade 3 compared to Grade 1 of Math, filled by guardian")
  #other than some that maybe dropped the course before getting the third grade, there appears to be a linear relationship, regardless of guardian. 

#----------------------------------------------------------------------------

#Scatterplots with faceting 
#Scatterplot of G3 as a Y axis and age with sex as a fill and school as a facet wrap 
ggplot(comb_data3, aes(x = age, y = G3.math, color = sex)) + geom_point() +
  geom_jitter(width = 0.35, opacity = 0.3) + labs(x = "Age", y = "Grade 3 of Math Course", title = "Grade 3 of math by age, filled by sex") + facet_wrap(~school)
#GP is much more variable in age, probably just due to having so many more students. There doesnt appear to be any significant differences in grade distributions between the schools here, or by sex in either plot.

#Scatterplot of G3 as a Y axis and G1 with guardian as a fill and school as a facet wrap
ggplot(comb_data3, aes(x = G1.math, y = G3.math, color = guardianF)) + geom_point() +
  geom_jitter(width = 0.35, opacity = 0.3) + labs(x = "Grade 1 of Math Course", y = "Grade 3 of Math Course", title = "Grade 3 compared to Grade 1 of Math, filled by guardian") + facet_wrap(~school)
  #It looks like both schools have about the same amounts of father vs mother vs other for guardian. There doesnt appear to be any significant differences in grade performance between schools; both follow a linear trend, the better a student did at grade 1 the better they did at grade 3, regardless of school or guardian situation. 

#----------------------------------------------------------------------------

#Scatterplots with faceting two variables 

#Scatterplot of G3 as a Y axis and age with sex as a fill and school and family size as a facet wrap 
ggplot(comb_data3, aes(x = age, y = G3.math, color = sex)) + geom_point() +
  geom_jitter(width = 0.35, opacity = 0.3) + labs(x = "Age", y = "Grade 3 of Math Course", title = "Grade 3 of math by age, filled by sex") + facet_wrap(school ~ famsizeF)
  #Both schools have slightly more students with families of more than 3 people. The students at MS that have less than or eqial to 3 family members look to have the best grade 3 scores out of all the groups- their lowest grade is around 7.5, whilst both family sizes of GP school and the other family size of MS have grade 0s. 

#Scatterplot of G3 as a Y axis and G1 with guardian as a fill and school and family size as a facet wrap
ggplot(comb_data3, aes(x = G1.math, y = G3.math, color = guardianF)) + geom_point() +
  geom_jitter(width = 0.35, opacity = 0.3) + labs(x = "Grade 1 of Math Course", y = "Grade 3 of Math Course", title = "Grade 3 compared to Grade 1 of Math, filled by guardian") + facet_wrap(school ~ famsizeF)
  #All four groups have a very similar linear trend, as G3 increases G1 increases, except that MS school's students with less than or equal to 3 family members dont have any 0 grades. Probably just an artifact 
```
